---
title: "Microbiota_Analysis"
author: "Adam Sorbie"
date: "26/10/2021"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis of Microbiota data

This document contains code to analyse the microbiota data generated in 
Sorbie, Jimenez and Benakis 2021 and is intended to act as template for other
stroke researchers to conduct their own analyses. 

```{r}
source("../../scripts/analysis_functions.R")
```

Firstly, we will read the ASV table(s) and metadata into R, using phyloseq. We have 
included the data used in this publication to demonstrate our analysis pipeline. 


```{r}
ps <- import_as_pseq(otu = "../../data/ASV_seqtab_tax.tab",
                mapping = "../../data/Metadata-16S-sequenced_wo_ctrls.txt",
                tree = "../../data/ASV_tree.tre")
```
Phyloseq is used as it provides a nice way of storing all the associated data in 
one object or class. As you can see, the metadata, tree and ASV table are all
combined here into ```ps```. Importantly, each part can also be accessed invidivually. 

## Data Normalisation 

Here will transform the data using minimum sum normalisation and relative abundance
for downstream analyses. 

### How does normalisation work? 

Minimum sum normalisation works by dividing the sum in each sample by the minimum 
sum across all samples, thereby standardising each sample to a fixed sum. 

Relative, transforms each sample into compositions, on a fixed scale of 0-100. 
```{r}
ps_norm <- transform(ps, transform = "mss")
ps_rel <- transform(ps, transform = "relative")
```

We can see how this works by looking at the column sums: 

```{r}
colSums(otu_table(ps_rel))
```

```{r}
colSums(otu_table(ps_norm))
```

## Alpha Diversity 

Here we will calculate three different measures of alpha-diversity:

* Species richness, or the number of observed species
  - This is calculated by 
* Shannon effective diversity, measuring 
* Faith's PD, measuring the phylogenetic component of alpha diversity 

The function ```calc_alpa``` wraps all of these calculations and only requires
the mss normalised phyloseq object as input, returning a dataframe with a column
for each above-mentioned dataframe. 

```{r}
alpha_div <- calc_alpha(ps_norm)
```

### How are alpha diversity metrics calculated? 

Richness here is calculated as the the total number of observed species greater 
than 0.5 mss normalised abundance. This threshold is used to exclude noise (see
Lagkouvardos et al 2017, PeerJ and Reitmeier et al 2021, ISME Communication 
for a more thorough explanation). 

Shannon effective diversity is calculated as the exponent of the Shannon index: 

$$H = -\sum_{i=1}^{R} p_i ln(p_i)$$
where *R* = richness, *p~i~* is the relative abundance of species *i* and *ln* is
the natural logarithm. 

This metric accounts for both the abundance and evenness of taxa. 

Lastly, Faith's alpha diversity is calulated as the sum of the branch lengths in the
phylogenetic tree. This method does not account for abundance, only measuring 
the phylogenetic diversity of a sample. 


Now we have calculated alpha diversity, we can merge this information with 
our metadata to generate a dataframe for plotting. 

```{r}
meta <-meta_to_df(ps_norm)
# merge alpha diversity df and metadata by rownames (0)
alpha_div_meta <- merge(alpha_div, meta, by=0)
```

### Plotting 

To plot the alpha diversity metrics we will use a boxplot with jittered points
layered on top. The function ```plot_boxplot```will do this for you, we just 
need to set some parameters first. 

Firstly, we will list the statistical comparisons we want to make, by creating 
a list of vectors. In this case we only have two groups, stroke and sham which 
can be written like: ```list(c("Stroke", "Sham"))```. If we had an extra group, 
for example, "Control", we would then have three comparisons to define like so: 
```list(c("Control", "Sham"), c("Control", "Stroke"), c("Sham", "Stroke"))```


Comparisons list  

```{r}
comps <- list(c("Stroke", "Sham"))
```


We can also specify the colours we want to use in our plots here by creating 
a named vector of colours. 

```{r}
colours <- c("Sham" = "dodgerblue2", "Stroke" = "firebrick1")
```

To generate the plot we need to provide the dataframe, the name of the column
containing the grouping variable (in this case simply "Group"), the name of the 
column containing the values to plot (Richness here). To colour by group we provide 
the column name of the grouping variable to ```fill_var```. We can then add
the list of comparisons, some x and y-axis, a title if desired and the plot colours.

In instances where the alphabetical order of your group levels does not match
the order you would like to plot them in, you can specify this here with
the ```group.order```parameter. 

#### Richness 
```{r}
plot_boxplot(df = alpha_div_meta, variable_col = "Group", value_col = "Richness", 
             fill_var = "Group", comparisons_list = comps, xlab = "Group", 
             ylab = "Richness", p_title = "Richness Stroke vs Sham", 
             col_palette = colours, group.order = c("Sham", "Stroke")) 
```
#### Shannon Effective 

```{r}
plot_boxplot(df = alpha_div_meta, variable_col = "Group", value_col = "Shannon.Effective", 
             fill_var = "Group", comparisons_list = comps, xlab = "Group", 
             ylab = "Shannon.Effective", p_title = "Shannon.Effective Stroke vs Sham", 
             col_palette = colours, group.order = c("Sham", "Stroke")) 
```

#### Faith's PD 

```{r}
plot_boxplot(df = alpha_div_meta, variable_col = "Group", value_col = "Faiths.PD", 
             fill_var = "Group", comparisons_list = comps, xlab = "Group", 
             ylab = "Faiths.PD", p_title = "Faith's PD Stroke vs Sham", 
             col_palette = colours, group.order = c("Sham", "Stroke")) 
```

Statistical significance is calculated internally in the ```plot_boxplot```
function using unpaired wilcoxon, thus currently this function is only
suitable when groups are independent. Note that if no significance is displayed 
on the plot, the the differences between groups were not statistically 
significant.

## Beta Diversity 

Here we will calculate beta-diversity based on Generalized unifrac distance and plot 
an ordination of this using Non-metric multidimensional scaling. 

The ```calc_betadiv``` function calculates a distance matrix, and an ordination 
of that matrix, returning both as a list. 

Various dissimilarity indices are available: 

* Bray-Curtis - A count-based dissimilarity metric (beta-diversity), 
based on the fraction of overabundant counts. 
* Unifrac - A phylogenetic beta-diversity metric measuring the fraction of 
unique branches in a phylogenetic tree.  
* Weighted Unifrac - An extensions of unifrac taking abundance into account in 
addition
* Generalized Unifrac - A further extension of unifrac, placing less weight on
highly-abundant lineages. 

Similarly, there are also various ordination options:

* NMDS (default) - Non-Metric Multidimensional Scaling. An ordination method 
which attempts to represent the dissimilarity between samples, as closely as 
possible in a low-dimensional space. 
* MDS/PCoA - Principal Coordinate analysis (also known as Metric Multidimensional Scaling). 
An ordination method which attempts to preserve distance between samples 
in a low dimensional Euclidean space.   

```{r}
betadiv <- calc_betadiv(ps_norm, dist = "gunifrac", ord_method = "NMDS")
```
To plot beta diversity a convenience function ```plot-beta_div``` is provided.
We just need to provide the phyloseq object, the ordination and distance matrix 
from the betadiv object above (This object is a list so we can access elements 
with $ notation and their name), a grouping variable and again the colours.

Since the ordination under the hood returns MDS as the column names, regardless 
of whether metric or non-metric dimensional scaling is used, we can adjust 
the axis labels by adding new x and y labels to our plot. All plotting functions
return ggplot2 objects which can be added to and further customised as desired. 

Within this function statistical testing of group separation is also
carried out using the adonis function of vegan. This function performs a Permutational
Multivariate Analysis of Variance or PERMANOVA test. The resulting R^2^ and p-value
are added to the plot in the bottom left. 

```{r}
plot_beta_div(ps_norm, ordination = betadiv$Ordination, 
              dist_matrix = betadiv$Distance_Matrix, group_variable = "Group", 
              cols = colours)+ 
  xlab("NMDS1") + 
  ylab("NMDS2")

```
## Differential abundance

The final step of this pipeline is to calculate differentially abundant 
taxa between conditions. 

This function performs the ancom-bc test, a compositionally aware differential
abundance method and returns significant results. As input, only the phyloseq
object and the column name of the grouping variable is required. 

```{r}
res_ancom <- ancom_da(ps_norm, "Group")
```
To visualise differentially abundant taxa, we provide a function which calculates
fold change of significant taxa from above and plots diverging dot plot coloured by 
group, providing a clear figure showing which taxa change between conditions. 

To this function, we need to provide the results of the ancom test above, 
an ordered vector of the group levels e.g. ```c("Sham", "Stroke")```. Additionally,
we can provide the group colours to make interpretation easier. 

```{r}
p <- plot_da(res_ancom, groups = c("Sham", "Stroke"), cols=colours)
```


