---
title: "Power_Analysis"
author: "Adam Sorbie"
date: "23/07/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Power Analysis for Microbiota studies 

```{r}
library(tidyverse)
library(powmic)
library(rMAGMA)
Sys.setlocale('LC_ALL','C')
```

## Power calculation 

This document contains code for the power analysis, associated with the manuscript:
Sorbie, Jimenez & Benakis, iScience 2022. 

To identify a suitable number of samples for our stroke microbiota study
we will use the package ```powmic``` (Chen, Bioinformatics 2020). To estimate 
parameters we will be using a similar dataset from Xu et al 2021, Gut. 

We will be using a subset of the mouse data here, looking at the day3 timepoint in
the colon as this is the closest to our own study design.  

```{r}
asv <- "../Xu_et_al_2021/data/longitudinal_cohort/ASV_seqtab.tab"
meta <- "../Xu_et_al_2021/data/longitudinal_cohort/meta_longitudinal.tab"

asvtab <- read_tsv(asv) %>% 
  column_to_rownames(var="#ASVId")
metadata <- read_tsv(meta) %>% 
  column_to_rownames(var="#SampleID")
```

```{r}
metadata_cn_d3 <- filter(metadata, Site == "Colon" & Timepoint == "3D")
asvtab_cn_d3 <- asvtab %>% dplyr::select(all_of(rownames(metadata_cn_d3)))
```


```{r}
nonzeros <- rowSums(asvtab_cn_d3 != 0)
asvtab_cn_d3 <- asvtab_cn_d3[nonzeros > 5,]
asvtab_mat <- as.matrix(asvtab_cn_d3)
```

Need to think about this a little bit here, 


```{r}
id <- match(colnames(asvtab_cn_d3),rownames(metadata_cn_d3))
group <- metadata_cn_d3$Group[id]
```

```{r}
data <- list()

data$counts <- asvtab_mat
data$designs <- numeric(length(group))
data$designs[which(group=='MCAO')]=1

d <- DGEList(counts=data$counts, group=data$designs)
d <- calcNormFactors(d)
d <- estimateCommonDisp(d)
d <- estimateTagwiseDisp(d)
fit <- exactTest(d)
res <- topTags(fit, n=nrow(data$counts))
res <- as.data.frame(res)
proportion <- sum(res$FDR<0.1)/nrow(data$counts)
proportion
lfc <- res$logFC[res$FDR<0.1]
hist(lfc)
```

```{r}
modelselect.AIC(t(asvtab_cn_d3))
```
```{r}
modelselect.vuong(t(asvtab_cn_d3))
```

```{r}
params <- list()

Sigma<- estSigma(t(asvtab_cn_d3),method='MAGMA',distrib='ZINB') #option for estimating microbial correlation structure
params$Sigma <- Sigma

out <- fit.ZINB(t(asvtab_cn_d3))
params$otu.mat <- t(asvtab_cn_d3)
params$mu <- out$mu
params$phi <- out$phi
params$p0 <- out$p0
```

```{r include=FALSE}
lmu0 <- log(params$mu)
lphi0 <- log(params$phi)
logp0 <- log(params$p0)
Sigma <- params$Sigma
distrib <- 'ZINB'
lfc.mu <- lfc
params.sim <- setParams.ZINB(nTaxa=1000,p.DA=0.05,Sigma = Sigma, lmu0=lmu0,
                             lphi0=lphi0,lp0 = logp0, lfc.mu=lfc)
powmic.out <- powmic(n1s=c(6,8, 10, 12), n2s=c(6, 8, 10, 12),
                  params=params.sim,distrib=distrib,nsims = 50)
```

```{r}
assess.out <- assess(powmic.out, alpha.type='fdr',alpha.level=0.1,stratify.type='prevalence')
sum.out <- summaryAssess(assess.out,assess.type='overall')
```

